tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-14 paper.tex

An important property of the succinct tree sequence ARG encoding
used by these population genetic simulators is that it allows us to
store vast genetic variation datasets very compactly. Instead of
storing the genetic sequence variation in the conventional
matrix-based form, we can store the underlying genealogical history
and mutations that ultimately \emph{result} in the observed
variation~\citep{kelleher2019}.
For example, a highly realistic simulation
of chromosome 9 for 1.4 million
French-Canadian individuals~\citep{anderson2023}
requires around 550GB of storage space in the
standard gzip-compressed VCF~\citep{danecek2011}
format ($\sim$280TB uncompressed).
The original simulation file representing
precisely the same genetic variation data, plus the
complete simulated ARG, requires only 1.36GB of space
in the compressed \texttt{tszip} format. Thus,
by encoding in terms of the underlying ARG,
in this case we achieve more than 400-fold compression,
and this highly unwieldy dataset
becomes straightforward to work with (we discuss
landmark dataset again in later sections).
Another substantial benefit of storing simulated datasets
in this ARG-based format is that many computations can
be expressed efficiently in terms of the underlying
ancestry~\citep{kelleher2016efficient,ralph2020}, without
needing to decode the actual variation data.

The first step of simulating data for evaluating GWAS methods,
generating realistic genetic variation datasets of millions
of individuals, is therefore at a high-level of sophistication
and seems sufficient to support ongoing research.
The second step, of generating the actual quantiative traits,
is currently lagging far behind.
Although there are multiple packages to simulate genotypic data
concurrently with quantitative traits~\citep{gaynor2021,haller2023},
and to simulate quantitative traits from a given set of genetic
sequences~\citep[e.g.][]{meyer2018,fernandes2020}, none can
handle the millions of individuals now required.
% TODO this needs another pass. We do need to address the Martin et
% al thing up front though, I think.
\citet{martin2017} simulated quantitative traits for
600,000 humans as part of their analysis of the portability
of polygenic risk scores across populations. This was
done by using custom scripts to generate traits for these
large ARGs output by the \texttt{msprime}
simulator~\cite{kelleher2016efficient}.
While the code provided by Martin et al.\ was sufficient
for their analysis, it is tightly coupled to the details of their study
and not documented or designed to be reused.
Simulation code can be subtle and difficult to
debug~\citep{ragsdale2020lessons}, and therefore there
is a critical need for a well-documented and
thoroughly tested means of simulating quantitative
traits directly from the efficient ARG representation used by
modern population genetic simulators.

In this paper we present \texttt{tstrait}, a new Python library that
efficiently simulates quantitative traits on an arbitrary ARG
in \texttt{tskit}~\citep{ralph2020} format.
Using the efficient data structures provided by \texttt{tskit},
\texttt{tstrait} can
quickly simulate quantitative traits for population-scale datasets,
with a very low memory overhead. This approach also provides a
richer model than existing methods,
allowing us to, for example, simulate traits for
ancestral individuals as well as contemporary samples.
The \texttt{tstrait} library also integrates well with the wider
PyData ecosystem of data-science tools,
using
\texttt{NumPy} arrays~\citep{numpy} and
\texttt{pandas} dataframes~\citep{pandas}
as input and output.

@@

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales.
Fig~\ref{fig:time}A shows how the simulation time scales
with the number of individuals using simple simulations
of humans using \texttt{stdpopsim}~\citep{adrion2020}.
[HERE]
The only published code to simulate quantitative traits
directly from ARGs is in \cite{martin2017}, which we compared with. Including
other tools that require the entire genome matrix is not practical, as the
simulation time will be dominated by processing the VCF file and it would not
be practically to analyze a large sample VCF file on a laptop PC. For example,
the Chromosome 9 French Canadian dataset takes approximately 280 terabytes in
the VCF format, but it is only 1.357 Gb in the \texttt{tskit} compressed
tree-sequence format and it only takes a few minutes to load and process by
using the tree sequence data structure.

>>>>

An important property of the succinct tree sequence ARG encoding
used by these population genetic simulators is that it allows us to
store enormous genetic variation datasets compactly. Instead of
storing the genetic sequence variation in the conventional
matrix-based form~\citep[e.g.\ VCF,][]{danecek2011},
we can store the underlying genealogical history
and mutations that ultimately \emph{result} in the observed
variation~\citep{kelleher2019}.
For example, a highly realistic simulation
of chromosome 9 for 1.4 million
French-Canadian individuals~\citep{anderson2023}
requires around 550GB of storage space in
gzip-compressed VCF ($\sim$280TB uncompressed).
The original simulation file representing
precisely the same genetic variation data, plus the
complete simulated ARG, requires only 1.36GB of space
in compressed \texttt{tszip} format. Thus,
by encoding in terms of the underlying ARG,
in this case we achieve more than 400-fold compression,
and this highly unwieldy dataset
becomes straightforward to work with (we discuss this
landmark dataset again in later sections).
Another substantial benefit of storing simulated datasets
in this ARG-based format is that many computations can
be expressed efficiently in terms of the underlying
ancestry~\citep{kelleher2016efficient,ralph2020}, without
needing to decode the actual variation data.

There are therefore sophisticated methods available for simulating
ARGs at the population scale, but there is currently no
easy way to generate quantiative traits based on such an ARG.
Existing methods to simulate quantitative traits from a given set of
genetic sequences in standard formats~\citep[e.g.][]{meyer2018,fernandes2020}
are overwhelmed by the scale of the data. For example,
just parsing the 280TB of VCF for the French-Canadian
dataset mentioned above would take many days of CPU time,
making quantitative trait simulation by this approach unfeasible.
% TODO this needs another pass. We do need to address the Martin et
% al thing up front though, I think.
In their analysis of the portability of polygenic risk
scores across populations, \citet{martin2017}
used efficient \texttt{tskit} APIs to simulate quantitative
traits for 600,000 humans directly from a simulated ARG.
While the code provided by Martin et al.\ works at this
very large scale, it is tightly coupled to the details of their study
and not designed to be reused.
Simulation code can be subtle and difficult to
debug~\citep{ragsdale2020lessons}, and there
is therefore a critical need for a well-documented and
thoroughly tested means of simulating quantitative
traits directly from an ARG.

In this paper we present \texttt{tstrait}, a new Python library that
efficiently simulates quantitative traits on an arbitrary ARG
in \texttt{tskit}~\citep{ralph2020} format.
\texttt{Tstrait} can
quickly simulate quantitative traits for population-scale datasets,
with a very low memory overhead. This approach also provides a
richer model than existing methods,
allowing us to, for example, simulate traits for
ancestral individuals as well as contemporary samples.
The \texttt{tstrait} library also integrates well with the wider
PyData ecosystem of data-science tools,
using
\texttt{NumPy} arrays~\citep{numpy} and
\texttt{pandas} dataframes~\citep{pandas}
as input and output.

@@

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales.
Fig~\ref{fig:time}A shows how the simulation time scales
with the number of individuals using simple simulations
of humans using \texttt{stdpopsim}~\citep{adrion2020}.
[HERE]
The only published code to simulate quantitative traits
directly from ARGs is in \cite{martin2017}, which we compared with. Including
other tools that require the entire genome matrix is not practical, as the
simulation time will be dominated by processing the VCF file and it would not
be practically to analyze a large sample VCF file on a laptop PC. For example,
the Chromosome 9 French Canadian dataset takes approximately 280 terabytes in
VCF, but it is only 1.357 Gb in the \texttt{tskit} compressed
tree-sequence format and it only takes a few minutes to load and process by
using the tree sequence data structure.
